---
title: "ggplot"
output: html_document
---
>The greatest value of a picture is when 
>it forces us to notice what we never expected to see.
>`r tufte::quote_footer('John Tukey')`

To put formatted R code in the text as such `ggplot()`, we use delimiters ` `` `. We can also evaluate the R code inline by using the format \"r+code" inside the delimitted text. 

We load the mpg data set.
```{r message=FALSE}
library(tidyverse)
mpg
```

A key idea behind the `ggolot` grammar of graphics lies in its syntax:
```{r eval=FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

After defining the data set with `data = <DATA>`, the additional layer is `<GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`. The type of plot if established by `<GEOM_FUNCTION>`. To establish the aesthetics of how the data is to appear under this plot type, the `mapping` argument maps aesthetic elements that are relevant to the plot type, to specific variables in the data. For example

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = manufacturer))
```

If on the other hand there is an aesthetic of the plot we wish to set which is a global feature and not tied to the presentation of a specific variable in the plot, it would appear as an argument outside of `mapping`. For example

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
It is also possible to simply embed the plot in the text as such (supressing the R code)
```{r echo=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

One way we have seen above to get information about additional variables in a plot is through aesthetics. Another approach is to break the plot into sub-plots, or *facets*. We can achieve this through `facet_wrap()` which is a layer in `ggplot` that takes a formula as an argument which must isolate categorical variables. Thus

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") +
  facet_wrap(~manufacturer, nrow = 1)
```

Or, alternatively

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") +
  facet_grid(.~manufacturer)
```

